Lossy compression of DNA sequencing quality data

### 2015-05-13 manuscript/q-compression.tex
Surprisingly large number of manuscript-improving edits.

Existing methods for compressing quality values mostly focus on
minimizing the loss of information with less emphasis on the effects
on subsequent analyses. In this paper, we evaluate several different
compression methods for quality values that compromise accuracy for
better storage efficiency, and their resulting impact on common
bioinformatic analyses using sequence read data.

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little negative effect on subsequent
analysis results. The three compression strategies in this study were
able to produce results similar to those obtained with uncompressed
quality values when performing quality control, genome assembly, and
alignment of short reads to a reference sequence.

@@

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2014rw,Giancarlo:2009fk,
  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
  janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}. Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective, some loss of information
is deemed acceptable if the inferences from analyses are relatively
unaffected. Sequence reads generated by instruments such as an
Illumina HiSeq, the focus of this research due to its preponderance in
genomics research, are characterized by having relatively few
insertion and deletion errors, but substitution (miscall) errors are
much more frequent and have context-specific patterns. These errors
are non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
and~\cite{janin2013adaptive} evaluated the effects of lossy
compression on identifying variants within a data set. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We build on these prior works, and assess the effects of quality
value information loss resulting from compression on subsequent
genomic analyses, including read preprocessing (filtering and
trimming), genome assembly, and read mapping.

@@

\cite{Wan:2012kq} provide three similar lossy compression strategies
based on binning the base error probability distribution: UniBinning,
Truncating, and LogBinning. UniBinning evenly splits the error
probability distribution into a user-defined number of partitions.
Truncating treats a user-defined number of highest quality values as a
single bin. LogBinning works similarly to UniBinning, except it uses
the \emph{log} of the error probability distribution, which
effectively bins the \textsc{ascii} quality values evenly. The 2-bin
encoding examined here can be viewed as something like a combination
of LogBinning and Truncating in that we place the highest quality
values (as defined above) of the log of the error probability
distribution into a single bin.

@@

Although we focus only on two bins in this work, a greater number of
bins can be used in practice, albeit with a higher cost in memory
requirement.  For example, an additional bin, thus giving bins of
good, bad, and intermediate, could be used for base qualities near the
border between bins in the 2-bin encoding. Additional bins beyond two,
and their resulting compressibility and effect on downstream analyses,
are not considered here.

@@

If quality values are modeled, compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. We model read
quality values as polynomial functions obtained with least-squares
fitting, as one approach to the compression of read quality values by
modeling. Although polynomial functions have significantly fewer
parameters than a read-length string of raw quality values (e.g., one
to six coefficients in our approach), the use of floating point
numbers to store coefficients greatly limits the compression potential
of the method. The coefficients could be represented with fewer bits,
perhaps with some loss of precision, but we have not considered
alternative representations in this study.

@@

\textit{Rhodobacter sphaeroides} 2.4.1, which are generated from a
fragment library (insert size of 180 nt; 2,050,868 paired-end reads)
and short-jump library (insert size of 3,500 nt; 2,050,868 reads; read
length 101 nt). The corresponding reference sequence was obtained from
the \textsc{ncbi} RefSeq database (NC\_007488.1, NC\_007489.1,
NC\_007490.1, NC\_007493.1, NC\_007494.1, NC\_009007.1, NC\_009008.1).

\textit{Homo sapiens} chromosome 14 data, which are generated from a
fragment library (insert size of 155 nt; 36,504,800 paired-end reads)
and short-jump library (insert sizes ranging from 2283-2803 nt;
22,669,408 reads; read length 101 nt). The corresponding reference
sequence was obtained from the \textsc{ncbi} RefSeq database
(NC\_000014.8).

\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
which are generated from a fragment library (insert size of 180 nt;
1,145,8940 paired-end reads; read length 151 nt). The corresponding
reference sequence was obtained from the \textsc{ncbi} RefSeq database
(NC\_000913.2).

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs after subsequent compression using the
Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, we use
bits/base-call, define it as the size of the compressed representation
of quality values (in bits) divided by the number of quality values
represented. As a measure of information loss we use mean squared
error (\textsc{mse}) as a loss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
quality values, $Q_i'$ is the compressed/decompressed quality value,
and $Q_i$ is the original quality value associated with position $i$.

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to
uncompressed data. Sickle starts at the ends of the read and uses a
sliding window ($0.1 \times$ the read length) to find locations where
the mean quality in the window falls below a given threshold (20 by
default). The sequence is then trimmed from this position to the read
end. If the trimmed sequence length is less than a certain threshold
(20 by default), then the sequence is discarded.

@@

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
different compressed representations (Figure
\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets;
the short-jump library results are available in Supplemental Table
1. Storing the uncompressed quality values requires 1 byte/base-call
because they are stored in \textsc{ascii} format, and the lossless
compression of each data set using \textsc{bz}ip2 ranges from
2.19--3.10 bits/base-call; these values provide a reference point
using a widely available compression program. Values from the same
class of compression method tend to cluster together across the
different data sets, with the 0-degree polynomial regression, profile
encodings, and \textsc{q}ual\textsc{c}omp have the lowest
bits/base-call.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more memory than the profile encoding methods for
only a $\sim$2--3$\times$ reduction in \textsc{mse}. When rate
parameter of \textsc{q}ual\textsc{c}omp is set to match the profile
encoding methods, \textsc{q}ual\textsc{c}omp performs slightly worse
in terms of \textsc{mse}. For the \textit{Rhodobacter sphaeroides}
fragment library, \textsc{q}ual\textsc{c}omp with rate set to 10
bits/read (0.099 bits/bp) has a \textsc{mse} of 17.29, whereas
256-profile encoding only requires 0.079 bits/bp and has a
\textsc{mse} of 11.85.

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression has the highest bits/base-call in the
\textit{Mus musculus} data set, and requires more storage than the
original \textsc{ascii} quality values. A 7th-degree polynomial
requires storing eight floating point values, resulting in 32 bytes
per read. The read length of the \textit{Mus musculus} data set is
only 26, so the 7th-degree polynomial regression is storing six more
bytes than necessary for lossless encoding of the quality data.

As an additional point of reference regarding compression
effectiveness, the recently published
Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
chromosome 21 data (read lengths ranging from 50-110 nt), and a mean
compression of 1.841 bits/base-call; both values are higher than all
of the profile results and for some results of other methods presented
here (Figure~\ref{fig:mse_vs_bpbp}).

@@

Preprocessing involves two steps: \emph{discarding} reads that are
deemed to be poor-quality overall, and \emph{trimming} the
poor-quality regions of the reads. After trimming, the majority of
compression methods retain more base-calls than the original quality
values (Figure \ref{fig:preprocessing}). In general, as a given
compression method increases in complexity --- i.e., as the number of
profiles, polynomial degrees, or rate increases --- the amount of
base-calls retained more closely approximates the number of base-calls
retained using the original quality values. The compression methods on
the \textit{Mus musculus} data set has the greatest proportion of
retained base-calls compared to the original quality values. The
\textit{Escherichia coli} MiSeq data set has the smallest range.

The 2-bin approach is the only compression method that results in a
higher number of trimmed base-calls compared to the original
uncompressed reads across all data sets. Sickle uses a sliding window
approach to smooth the read quality values before it trims. In the
2-bin approach, there is an uneven distribution of values per bin. In
other words, \emph{bad} quality values may range from 0--33, whereas
\emph{good} values may only range from 34--40. Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
trimmed bases.

@@

It is important to highlight that even though a compression method may
result in the same number of trimmed base-calls as the uncompressed
quality values, it does not mean the \emph{same} base-calls were
retained.  For example, pre-processing of the 1st-degree and
5th-degree polynomial regression model compressed reads of the
\textit{Rhodobacter sphaeroides} fragment library retain approximately
the same number of base-calls. However, if we examine the specific
reads discarded, the 5th-degree model discards approximately
two-thirds less reads than the 1st-degree model (4,640 and 12,066
reads, respectively) (Supplemental Table 2).

@@

No compression method resulted in the uniformly best assembly of the
\textit{Rhodobacter sphaeroides} data set in all metrics (Table
\ref{fig:assembly_ranks}). Among the compression methods, the profile
encoding performed best, polynomial regression modeling performed
worst, and other methods were intermediate
(Fig.~\ref{fig:assembly_ranks}).

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with unmodified quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbp (see
Supplemental Table 3). Despite the similar contiguity statistics, the
different compression methods vary noticeably in the amount of
\textsc{snp}s. The 2-bin and profile methods exhibited the fewest
\textsc{snp}s compared to the reference genome, thus outperforming the
assembly using the original quality values in this characteristic. A
more in-depth evaluation is needed to determine whether these
compression methods are missing actual \textsc{snp}s.

It is important to highlight that using uncompressed reads does not
produce the best assembly in terms of any of the metrics. The
uncompressed reads score worse than the top overall assembly
(256-profile encoding) in terms of assembled bases, missing reference
bases, N50, \textsc{snp}s, indels $>$5bp, and relocations. The
assembly using the original uncompressed reads has an error rate of
$\sim$8.75 errors/100 kb of assembled sequence, and the 256-profile
encoding has an error rate of $\sim$8.02 errors/100 kbp (Supplemental
Table 3).

In general, the greater the polynomial degree, the better the overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poorly in all metrics except assembled
bases. One explanation is that the high-error portions of reads are
being marked as high quality, so \textsc{allpaths-lg} is unable to
trim or error-correct the sequences. Assembled sequences that overlap
may be unable to align across the errors at the end of the reads,
artificially inflating the assembled genome size.

Among the different \textsc{q}ual\textsc{c}omp rate parameters, the 10
bits/read rate ranked highest overall, outperforming the other rate
parameters in terms of corrected N50, fewest missing reference bases,
\textsc{snp}s, and indels $>$5bp. With the exception of the 6
bits/read rate, the assemblies decrease in rank with the increase in
the rate parameter in terms of corrected N50, and fewest missing
reference bases. This trend runs counter to the decrease in
\textsc{mse} of the different rates.

@@

Certain short read alignment tools use the quality value information
when finding potential alignments. Bowtie2 (version 2.2.3) was used to
evaluate the different decompressed \textsc{fastq} files. Bowtie2 uses
quality values written in the \textsc{fastq} files when mapping reads
against a reference genome. The original uncompressed and decompressed
\textsc{fastq} files were mapped with Bowtie2 against the
\textit{Rhodobacter sphaeroides} reference genome. The generated
\textsc{sam} file for each compression approach was compared with the
uncompressed \textsc{sam} file. The total, shared and unique
proportional numbers of mapped reads are calculated with respect to
the uncompressed \textsc{sam} matches as shown in Table
\ref{tab:aligner}. Additionally, to monitor the effect of quality
values on mapping in general, Bowtie2 was adjusted so that the maximum
and minimum mismatch penalty were equivalent to maximum and minimum
quality scores (with parameters: --mp 6,6 and --mp 2,2 respectively).

We evaluate the compression methods using two approaches. In the first
approach, we order the compression methods based on how similar the
alignment results are using the original uncompressed quality values
--- i.e., the amount of reads aligned by both the uncompressed and
compressed methods plus the amount of reads unaligned by both methods
minus the amount of reads uniquely aligned by the uncompressed and
compression method. In the second approach, we order the compression
methods by total proportion of aligned reads.

@@

Most of the compression methods did not vary greatly in terms of the
number of reads that were mapped \emph{only} by the compression
method; however, there is a sizable difference in the amount of reads
that are originally mapped, but unmapped by the compressed methods.
\textsc{q}ual\textsc{c}omp with rate 100 bits/read results in the
fewest missing original read alignments (159). Increasing the
regression model polynomial degree results in a decreasing amount of
reads that are originally mapped, but unmapped by the regression model
(40,931 to 1,134 reads for 0-degree and 7th-degree,
respectively). There is no such trend for reads that are mapped only
by the regression model.

@@

Downstream applications dictate which lossy compression methods are
acceptable. As noted above, 2-bin encoding is able to retain
information about low-quality base-calls that occur within stretches
of high-quality base-calls, whereas the polynomial regression methods
tend to smooth these quality values. In addition to aligning fewer
reads than the 2-bin encoding, the polynomial regression methods may
also be hindering downstream tools that use the alignment results. A
\textsc{snp} detection pipeline will commonly use the quality values
to record if a potential difference from the reference is the result
of genetic variation or sequencing error. Erroneously marking a base
as higher quality may lead to the base being incorrectly labeled as a
\textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly, and 7th-degree polynomial
regression modeling outperforms it in nearly all metrics. Profiling
performs quite well in all use cases.

@@

We have examined several simple and general approaches for lossy
compression of read quality values and their effect on subsequent
analyses. Although most previous examinations of lossy compression
focused primarily on information loss and compression effectiveness,
we show that common bioinformatic analyses are robust to substantial
loss of information in quality values. Some methods, such as
profiling, have yield excellent compression and provide excellent
results in subsequent analyses, and in some cases surpassing results
from original uncompressed data.

@@

\section*{Acknowledgement}
This project was initiated at the 2014 Bioinformatics Exchange for
Students and Teachers (\textsc{best}) Summer School, which was funded
by the offices of the Dean of The Graduate School, University of
Maryland, and the Rektor of the University of T\"{u}bingen. Additional
funding included an International Graduate Research Fellowship from
The Graduate School, University of Maryland, to CMH, a Global
Partnerships-Faculty Travel Grant from the Office of International
Affairs to MPC, and other funding from the University of T\"{u}bingen.

>>>>

Existing methods for compressing quality values mostly focus on
minimizing the loss of information with less emphasis on the effects
on subsequent analyses. In this paper, we evaluate several different
compression methods for quality values that compromise accuracy for
better storage efficiency, and assess the resulting impacts on common
bioinformatic analyses using sequence read data.

\section{Results:}
Lossy compression of quality information can greatly decrease storage
and memory requirements with little negative effect on subsequent
analysis results. The three compression strategies in this study were
able to produce results similar to those obtained with the original
uncompressed quality values when performing quality control, genome
assembly, and alignment of short reads to a reference sequence.

@@

Previous literature on sequence data compression has largely focused
on lossless compression of base calls~\cite[reviewed
  in][]{Deorowicz:2013hq,Giancarlo:2009fk,Giancarlo:2014rw,
  Nalbantoglu:2010uq,Zhu:2013qr}, although some recent work has
addressed compression of quality
values~\cite[e.g.,][]{asnani2012lossy,Canovas:2014fr,Hach:2012ys,
  janin2013adaptive,Kozanitis:2011kl,Ochoa:2013rt,Tembe:2010ys,
  Wan:2012kq,DBLP:conf/recomb/YuYB14,zhou2014compression}. Among the
several challenges for compression of read data is dealing with
different error profiles resulting from differences in underlying
chemistries, signal detection and processing mechanisms, inherent
biases, and other idiosyncratic properties of distinct high-throughput
sequencing technologies.

Although we recognize the need for lossless compression for some
purposes and contexts (e.g., archiving, provenance), our perspective
is largely pragmatic with a focus on the use of quality values in
subsequent analyses. From this perspective, some loss of information
is deemed acceptable if the inferences from analyses are relatively
unaffected. Sequence reads generated by instruments such as an
Illumina HiSeq, the focus of this research due to its preponderance in
genomics, are characterized by having relatively few insertion and
deletion errors, but substitution (miscall) errors are much more
frequent and have context-specific patterns. These errors are
non-uniformly distributed over the read length (e.g., error rates
increase up to $\sim$16$\times$ at the 3$^{\prime}$ end, and 32.8 --
67.9\% of reads have low-quality tails at the 3$^{\prime}$
end~\citep{Minoche:2011km}). Recognizing these properties of Illumina
sequence reads motivates our exploration of three general classes of
lossy compression methods -- binning, modeling, and profiling -- and
our consideration of an exemplar of each class.~\cite{Canovas:2014fr}
and~\cite{janin2013adaptive} evaluated the effects of lossy
compression on identifying variants within a data set. Here we
describe our research investigating lossy compression of sequence read
quality values, specifically those associated with Illumina
instruments, with the objective to provide some perspective on several
strategies rather than to develop robust high-quality software for
use. We assess the effects of quality value information loss resulting
from compression on subsequent genomic analyses, including read
preprocessing (filtering and trimming), genome assembly, and read
mapping.

@@

\cite{Wan:2012kq} provide three similar lossy compression strategies
based on binning the base error probability distribution: UniBinning,
Truncating, and LogBinning. UniBinning evenly splits the error
probability distribution into a user-defined number of partitions.
Truncating treats a user-defined number of highest quality values as a
single bin. LogBinning works similarly to UniBinning, except it uses
the \emph{log} of the error probability distribution, which
effectively bins the \textsc{ascii} quality values evenly. The 2-bin
encoding examined here can be viewed as something like a combination
of LogBinning and Truncating in that we place the highest quality
values (as defined above) from the distribution of log error
probability values into a single bin.

@@

Although we focus only on two bins in this work, a greater number of
bins can be used in practice, albeit with a higher cost in memory
usage. For example, an additional bin, thus giving bins of good, bad,
and intermediate, could be used for quality values near the border
between bins in the 2-bin encoding. Additional bins beyond two, and
their resulting compressibility and effect on downstream analyses, are
not considered here.

@@

If quality values are modeled, compression is conceivably possible by
replacing the original quality values by a representation of the
model. For example, quality values can be conceptualized as bivariate
data with the ordered nucleotides (1 to read length) representing the
abscissa, and quality values representing the ordinate. We model read
quality values as polynomial functions obtained with least-squares
fitting, as one approach to the compression of read quality values by
modeling. Although polynomial functions have significantly fewer
parameters than a read-length string of raw quality values (e.g., one
to six coefficients in our approach), the use of floating point
numbers to store coefficients greatly limits the compression potential
of the models. The coefficients could be represented with fewer bits,
perhaps with some loss of precision, but we have not considered
alternative representations in this study.

@@

\textit{Rhodobacter sphaeroides} 2.4.1, which are generated from a
fragment library (insert size of 180 bp; 2,050,868 paired-end reads;
read length 101 nt) and short-jump library (insert size of 3,500 bp;
2,050,868 reads; read length 101 nt). The corresponding reference
sequence was obtained from the \textsc{ncbi} RefSeq database
(NC\_007488.1, NC\_007489.1, NC\_007490.1, NC\_007493.1, NC\_007494.1,
NC\_009007.1, NC\_009008.1).

\textit{Homo sapiens} chromosome 14 data, which are generated from a
fragment library (insert size of 155 bp; 36,504,800 paired-end reads)
and short-jump library (insert sizes ranging from 2283-2803 bp;
22,669,408 reads; read length 101 nt). The corresponding reference
sequence was obtained from the \textsc{ncbi} RefSeq database
(NC\_000014.8).

\textit{Escherichia coli} str. K-12 MG1655 MiSeq data was downloaded
from \url{http://www.illumina.com/systems/miseq/scientific_data.html},
which are generated from a fragment library (insert size of 180 bp;
1,145,8940 paired-end reads; read length 151 nt). The corresponding
reference sequence was obtained from the \textsc{ncbi} RefSeq database
(NC\_000913.2).

@@

As a measure of compression effectiveness, which reflects the sum of
fixed and marginal costs after subsequent compression using the
Burrows-Wheeler algorithm~\citep{bwt} via \textsc{bz}ip2, we use
bits/base-call, defined as the size of the compressed representation
of quality values (in bits) divided by the number of quality values
represented. As a measure of information loss we use mean squared
error (\textsc{mse}) as a loss function, and define it as
$\frac{1}{n}\sum_{i=1}^{n}{(Q_i'-Q_i)^2}$, where $n$ is the number of
quality values, $Q_i'$ is the compressed/decompressed quality value,
and $Q_i$ is the original quality value associated with position $i$.

We evaluate effects of information loss from quality value compression
on quality control steps of trimming and read filtering, which were
performed using Sickle~\citep{sickle}, and make comparison to original
uncompressed data. Sickle starts at the ends of the read and uses a
sliding window (0.1 $\times$ the read length) to find locations where
the mean quality in the window falls below a given threshold (20 by
default). The sequence is then trimmed from this position to the read
end. If the trimmed sequence length is less than a certain threshold
(20 by default), then the sequence is discarded.

@@

We compare the quality-value \textsc{mse} versus bits/base-call of the
\textit{Rhodobacter sphaeroides}, \textit{Homo sapiens},
\textit{Escherichia coli}, and \textit{Mus musculus} data sets for
quality values resulting from the compression methods examined (Figure
\ref{fig:mse_vs_bpbp}). Here we present the fragment libraries for the
\textit{Rhodobacter sphaeroides}, and \textit{Homo sapiens} data sets;
the corresponding short-jump library results are available in
Supplemental Table 1. Storing the original uncompressed quality values
requires 1 byte/base-call because they are stored in \textsc{ascii}
format, and the lossless compression of each original data set using
\textsc{bz}ip2 ranges from 2.19--3.10 bits/base-call; these values
provide a reference point using a widely available general compression
program. Values from the same class of compression method tend to
cluster together across the different data sets, with the 0-degree
polynomial regression, profile encodings, and
\textsc{q}ual\textsc{c}omp yielding the lowest bits/base-call values.

\textsc{q}ual\textsc{c}omp with the rate parameter set to 100
bits/read has the lowest \textsc{mse}, but requires
$\sim$10--15$\times$ more memory than the profile encoding methods for
only a $\sim$2--3$\times$ reduction in error. When the rate parameter
of \textsc{q}ual\textsc{c}omp is set to match the profile encoding
methods, \textsc{q}ual\textsc{c}omp performs slightly worse in terms
of \textsc{mse}. For the \textit{Rhodobacter sphaeroides} fragment
library, \textsc{q}ual\textsc{c}omp with rate set to 10 bits/read
(0.099 bits/bp) has a \textsc{mse} of 17.29, whereas 256-profile
encoding only requires 0.079 bits/bp and has a \textsc{mse} of 11.85.

As the degree of the polynomial increases, the bits/base-call increase
and the \textsc{mse} decreases at an exponential rate. The 7th-degree
polynomial regression results in the largest bits/base-call, as it
requires 32 bytes per read before subsequent compression with
\textsc{bz}ip2. For the \textit{Mus musculus} data set, which has read
lengths of only 36 nt, the memory required for storage, even after
subsequent compression with \textsc{bz}ip2 exceeds that for the
original \textsc{ascii} quality values.

As an additional point of reference regarding compression
effectiveness, the recently published
Read-Quality-Sparsifier~\citep{DBLP:conf/recomb/YuYB14} achieved
best-case compression of 0.254 bits/base-call on \textit{Homo sapiens}
chromosome 21 data (read lengths ranging from 50--110 nt), and a mean
compression of 1.841 bits/base-call; both values are larger than all
of the profile results and for some results of other methods presented
here (Figure~\ref{fig:mse_vs_bpbp}).

@@

Preprocessing involves two steps: \emph{discarding} reads that are
deemed to be poor-quality overall, and \emph{trimming} the
poor-quality regions of the reads. After trimming, the majority of
compression methods resulted in retaining more base-calls than the
original quality values (Figure \ref{fig:preprocessing}). In general,
as a given compression method increases in complexity --- i.e., as the
number of profiles, polynomial degrees, or rate increases --- the
amount of base-calls retained more closely approximates the number of
base-calls retained using the original quality values. The compression
methods on the \textit{Mus musculus} data set has the greatest
proportion of retained base-calls compared to the original quality
values. The \textit{Escherichia coli} MiSeq data set has the smallest
range.

The 2-bin approach is the only compression method that results in a
larger number of trimmed base-calls compared to the original
uncompressed reads across all data sets. Sickle uses a sliding window
approach to smooth the read quality values before it trims. In the
2-bin approach, there is an uneven distribution of values per bin. In
other words, \emph{bad} quality values may range from 0--33, whereas
\emph{good} values may only range from 34--40. Thus, mid-range quality
values that are above the threshold (20 by default) are set below the
quality threshold when compressed, resulting in an increased number of
trimmed bases.

@@

It is important to highlight that even though a compression method may
result in the same number of trimmed base-calls as the uncompressed
quality values, it does not mean the \emph{same} base-calls were
retained.  For example, pre-processing of the 1st-degree and
5th-degree polynomial regression model compressed reads of the
\textit{Rhodobacter sphaeroides} fragment library retain approximately
the same number of base-calls. However, if we examine the specific
reads discarded, the 5th-degree model discards approximately
two-thirds less reads than the 1st-degree model (4,640 and 12,066
reads, respectively; Supplemental Table 2), which means there are
differences in the base-calls trimmed.

@@

No compression method resulted in the uniformly best assembly of the
\textit{Rhodobacter sphaeroides} data set in all metrics, although
several methods resulted in better assemblies than the original
uncompressed data (Table \ref{fig:assembly_ranks}). Among the
compression methods, the profile encoding performed best, polynomial
regression modeling performed worst, and other methods were
intermediate (Fig.~\ref{fig:assembly_ranks}).

The lossy compression methods largely preserve the contiguity found in
the assembly produced using the reads with the original quality
values. All compression methods other than 0-degree polynomial
regression produce an N50 ranging from 3.17--3.22 Mbp (see
Supplemental Table 3). Despite the similar contiguity statistics, the
different compression methods vary markedly in the number of
\textsc{snp}s. The 2-bin and profile methods exhibited the fewest
\textsc{snp}s compared to the reference genome, thus outperforming the
assembly using the original quality values in this characteristic. A
more in-depth evaluation is needed to determine whether these
compression methods are missing actual \textsc{snp}s.

It is important to highlight that using the original uncompressed
quality values does not produce the best assembly in terms of any of
the metrics. The assembly based on the original uncompressed quality
values score worse than the top overall assembly (256-profile
encoding) for number of assembled bases, missing reference bases, N50,
\textsc{snp}s, indels $>$5bp, and relocations. The assembly using the
original uncompressed quality values has an error rate of $\sim$8.75
errors/100 kb of assembled sequence, and the 256-profile encoding has
an error rate of $\sim$8.02 errors/100 kbp (Supplemental Table 3).

In general, the greater the polynomial degree, the better the overall
assembly; however, the 5th-degree polynomial regression performs
slightly worse than the 3rd-degree polynomial. The respective ranks in
terms of N50 and relocations are fairly distant, which lowers the
overall ranking of the 5th-degree polynomial slightly below that for
the 3rd-degree polynomial model. The 1st- and 0-degree polynomial
regression methods perform poorly in all metrics except assembled
bases. One explanation for this observation is that the high-error
portions of reads are being marked as high quality, so
\textsc{allpaths-lg} is unable to trim or error-correct the
reads. Assembled sequences that overlap may be unable to align across
the errors at the end of the reads, artificially inflating the
assembly size.

Among the different \textsc{q}ual\textsc{c}omp rate parameters, the 10
bits/read rate ranked highest overall, outperforming the other rate
parameters in terms of corrected N50, fewest missing reference bases,
\textsc{snp}s, and indels $>$5bp. With the exception of the 6
bits/read rate, the assemblies decrease in rank with the increase in
the rate parameter for corrected N50, and fewest missing reference
bases. This trend runs counter to the decrease in \textsc{mse} of the
different rates.

@@

Some short read alignment tools, such as Bowtie2 (version 2.2.3),
which was used here, utilize quality value information when evaluating
potential alignments. The reads from with original uncompressed and
compressed/decompressed quality vlaues were mapped with Bowtie2 to the
\textit{Rhodobacter sphaeroides} reference genome. The total, shared,
and unique proportion of mapped reads are calculated with respect to
the results for the the original uncompressed quality values as shown
in Table \ref{tab:aligner}. Additionally, to assess the effect of
quality values on mapping in general, Bowtie2 was adjusted so that the
maximum and minimum mismatch penalty were equivalent to maximum and
minimum quality scores (with parameters: --mp 6,6 and --mp 2,2
respectively).

We evaluate the compression methods using two approaches. In the first
approach, we order the compression methods based on how similar the
alignment results are using the original uncompressed quality values
--- i.e., the number of reads aligned using the original uncompressed
quality values and quality values resulting from compression, plus the
number of reads unaligned using the original uncompressed quality
values and quality values resulting from compression minus the amount
of reads uniquely aligned using the original uncompressed quality
values and quality values resulting from compression. In the second
approach, we order the compression methods by total proportion of
aligned reads.

@@

Most of the compression methods did not vary greatly in terms of the
number of reads that were mapped \emph{only} using quality values
resulting from compression; however, there is a sizable difference in
the amount of reads that are originally mapped, but unmapped by the
compressed methods.  \textsc{q}ual\textsc{c}omp with rate 100
bits/read results in the fewest missing original read alignments
(159). Increasing the regression model polynomial degree results in a
decreasing amount of reads that are originally mapped, but unmapped by
the regression model (40,931 and 1,134 reads for 0-degree and
7th-degree, respectively). There is no such trend for reads that are
mapped only by the regression model.

@@

Downstream applications dictate which lossy compression methods are
acceptable. As noted above, 2-bin encoding is able to retain
information about low-quality base-calls that occur within stretches
of high-quality base-calls, whereas the polynomial regression methods
tend to smooth these quality values. In addition to aligning fewer
reads than the 2-bin encoding, the polynomial regression methods may
also be hindering downstream tools that use the alignment results. A
\textsc{snp} detection pipeline will commonly use the quality values
to record if a potential difference from the reference is the result
of genetic variation or sequencing error. Erroneously marking a base
as higher quality may lead to the base being incorrectly labeled as a
\textsc{snp}. 2-bin encoding is acceptable for alignment; however, it
performs quite poorly in terms of assembly, and 7th-degree polynomial
regression modeling outperforms it in nearly all metrics. Profiling
performs quite well in all use cases, and has the best performance in
terms of bits/base-call versus error.

@@

We have examined several simple and general approaches for lossy
compression of read quality values and their effect on subsequent
analyses. Although most previous examinations of lossy compression
focused primarily on information loss and compression effectiveness,
we show that common bioinformatic analyses are robust to substantial
loss of information in quality values. Some methods, such as
profiling, have yield excellent compression and provide excellent
results in subsequent analyses, and in some cases surpass performance
of the original uncompressed data.

@@

\section*{Acknowledgement}
This project was initiated at the 2014 Bioinformatics Exchange for
Students and Teachers (\textsc{best}) Summer School, which was funded
by the offices of the Dean of The Graduate School, University of
Maryland, and the Rektor of the University of T\"{u}bingen. Additional
funding included an International Graduate Research Fellowship from
The Graduate School, University of Maryland, to CMH, a Global
Partnerships-Faculty Travel Grant from the Office of International
Affairs to MPC, and funding from the University of T\"{u}bingen.
